### 2018-02-09 submission/manuscript.Rmd
update with ALL genera model and OR

***Mouth-Associated Genera are Associated with an Increased OR of Tumor:***
Next, we asked if being higher than the median relative abundance, for any specific genera, resulted in an altered OR for adenoma or carcinoma, in stool and colon tissue, due to our previous observations of small increases in OR using OTU richness and Shannon diversity. To investigate this we analyzed all common genera across each study, in colon tissue or stool, and assessed whether a relative abundance higher than the median results in an increase or decrease in RR. Mouth-associated genera were commonly found in the top 5 genera associated with an increased OR of having an adenoma (*Porphyromonas* [Figure 3A] and *Rothia* [Figure 3C]) and carcinoma (*Fusobacterium*, *Parvimonas*, *Porphyromonas*, and *Peptostreptococcus* [Figure 3B] and *Fusobacterium* [Figure 3D]) for both stool and colon tissue samples. Conversely, genera commonly associated with a normal gastrointestinal tract were correlated with a decreased OR for both adenoma and carcinoma for both stool and colon tissue samples [Figure 3]. Even though mouth-associated genera were identified across disease stage, there was little direct overlap of the top 5 increased or decreased OR genera between both stages and sample site.

When observing ORs with a P-value less than 0.05 there was almost no overlap between genera from stool or colon tissue and when they were similar the RR was in opposite directions (e.g. *Lactococcus*) [Table S8 & S9]. Many of the genera that had ORs with a P-value under 0.05 for colon tissue are also highly prevalent in contamination, specifically, *Novosphingobium*, *Pseudomonas*, and *Achromobacter* [Figure 3 & Table S8-S9]. For carcinoma stage of disease, certain mouth-associated genera (*Fusobacterium*, *Parvimonas*) had an increased OR for both colon tissue and stool samples [Table S10 & S11]. Finally, these data suggest that the most significantly increased OR genera for tissue was *Lepttrichia* while in stool it was *Peptostreptococcus* [Table S10 & S11].

***Select Community Models can Recapitulate Whole Community Models:***
Since specific genera increased OR for carcinoma over diversity metrics we assessed whether the bacterial community was better at classifying disease versus only a select group of genera. We selected these genera based on their OR and P-value significance and used two approaches to test this question.
The first approach used genus level data and tested for differences in AUC between all genera and selected genera. A single study was used for training the model prior to testing on all other studies and this was repeated for every study in the meta-analysis. The second approach used OTU level data and tested for a generalized decrease in the 10-fold cross validation (CV) model AUC which is a common approach used to guard against over-fitting. This was applied across study and the AUC of the all OTUs model was compared against the model that used only OTUs that taxonomically classified to selected genera.

For the first approach using the genera-based models, the training set median AUC for model classification was similar for both the select genera and full genera models, for both tissue and stool studies [Figure S2-S3]. When analyzing the tests sets that were comprised of genera data from other studies, both models had a similar ability to detect individuals with adenomas or carcinomas, with the select genera models performing better in some instances [Figure S4-S6]. Conversely, the second approach that used OTU-based models showed a slight decrease in median AUC between the full and select models, with one exception to this generalization being the carcinoma models for matched colon tissue [Figure 4 & 5].

@@

**Figure S1: Relative Risk for Adenoma or Carcinoma based on Bacterial Community Metrics in Tissue.** A) Community-based relative risk for adenoma. B) Community-based relative risk for carcinoma. Colors represent the different variable regions used within the respective study.

>>>>

***Mouth-Associated Genera are Associated with an Increased OR of Tumor:***
Next, we asked if being higher than the median relative abundance, for any specific genera, resulted in an altered OR for adenoma or carcinoma, in stool and colon tissue, due to our previous observations of small increases in OR using OTU richness and Shannon diversity. To investigate this we analyzed all common genera across each study, in colon tissue or stool, and assessed whether a relative abundance higher than the median results in an increase or decrease in OR. Mouth-associated genera were commonly found in the top 5 genera associated with an increased OR of having an adenoma (*Porphyromonas* [Figure 3A] and *Rothia* [Figure 3C]) and carcinoma (*Fusobacterium*, *Parvimonas*, *Porphyromonas*, and *Peptostreptococcus* [Figure 3B] and *Fusobacterium* and *Parvimonas* [Figure 3D]) for both stool and colon tissue samples. Conversely, genera commonly associated with the gastrointestinal tract were correlated with a decreased OR for both adenoma and carcinoma for both stool and colon tissue samples [Figure 3]. Even though mouth-associated genera were identified across disease stage, there was little direct overlap of the top 5 increased or decreased OR genera between both stages and sample site.

When observing ORs for adenoma between genera from stool or colon tissue with a P-value less than 0.05 there was almost no overlap and when they were similar the OR was in opposite directions (e.g. *Lactococcus*) [Table S8 & S9]. Many of the adenoma associated genera ORs with a P-value under 0.05 for colon tissue are also highly prevalent in contamination, specifically, *Novosphingobium*, *Pseudomonas*, and *Achromobacter* [Figure 3 & Table S8-S9]. For carcinoma stage of disease, certain mouth-associated genera (*Fusobacterium*, *Parvimonas*) had an increased OR for both colon tissue and stool samples [Table S10 & S11]. The genera with the highest increased OR for carcinoma in tissue was *Lepttrichia* while in stool it was *Peptostreptococcus* [Table S10 & S11].

***Select Community Models can Recapitulate Whole Community Models:***
Since specific genera increased the OR for carcinoma over diversity metrics we assessed whether the bacterial community was better at classifying disease versus only a select group of genera. We selected these genera based on their OR and P-value significance and used two approaches to test this question.
The first approach used genus level data and tested for differences in AUC between all genera and selected genera. A single study was used for training the model prior to testing on all other studies and this was repeated for every study in the meta-analysis. The second approach used OTU level data and tested for a generalized decrease in the 10-fold cross validation (CV) model AUC which is a common approach used to guard against over-fitting. This was applied across study and the AUC of the all OTUs model was compared against the model that used only OTUs that taxonomically classified to selected genera.

For the first approach using the genera-based models, the training set median AUC for model classification was similar for both the full and select genera models, for both tissue and stool studies [Figure S2-S3]. When analyzing the tests sets that were comprised of genera data from other studies, both models had a similar ability to detect individuals with adenomas or carcinomas, with the select genera models performing better in some instances [Figure S4-S6]. Conversely, the second approach that used OTU-based models showed a slight decrease in median AUC between the full and select models [Figure 4 & 5].

@@

**Figure S1: Odds Ratio for Adenoma or Carcinoma based on Bacterial Community Metrics in Tissue.** A) Community-based odds ratio for adenoma. B) Community-based odds ratio for carcinoma. Colors represent the different variable regions used within the respective study.

### 2018-02-09 submission/manuscript.tex
update with ALL genera model and OR

\textbf{\emph{Mouth-Associated Genera are Associated with an Increased
OR of Tumor:}} Next, we asked if being higher than the median relative
abundance, for any specific genera, resulted in an altered OR for
adenoma or carcinoma, in stool and colon tissue, due to our previous
observations of small increases in OR using OTU richness and Shannon
diversity. To investigate this we analyzed all common genera across each
study, in colon tissue or stool, and assessed whether a relative
abundance higher than the median results in an increase or decrease in
RR. Mouth-associated genera were commonly found in the top 5 genera
associated with an increased OR of having an adenoma
(\emph{Porphyromonas} {[}Figure 3A{]} and \emph{Rothia} {[}Figure 3C{]})
and carcinoma (\emph{Fusobacterium}, \emph{Parvimonas},
\emph{Porphyromonas}, and \emph{Peptostreptococcus} {[}Figure 3B{]} and
\emph{Fusobacterium} {[}Figure 3D{]}) for both stool and colon tissue
samples. Conversely, genera commonly associated with a normal
gastrointestinal tract were correlated with a decreased OR for both
adenoma and carcinoma for both stool and colon tissue samples {[}Figure
3{]}. Even though mouth-associated genera were identified across disease
stage, there was little direct overlap of the top 5 increased or
decreased OR genera between both stages and sample site.

When observing ORs with a P-value less than 0.05 there was almost no
overlap between genera from stool or colon tissue and when they were
similar the RR was in opposite directions (e.g. \emph{Lactococcus})
{[}Table S8 \& S9{]}. Many of the genera that had ORs with a P-value
under 0.05 for colon tissue are also highly prevalent in contamination,
specifically, \emph{Novosphingobium}, \emph{Pseudomonas}, and
\emph{Achromobacter} {[}Figure 3 \& Table S8-S9{]}. For carcinoma stage
of disease, certain mouth-associated genera (\emph{Fusobacterium},
\emph{Parvimonas}) had an increased OR for both colon tissue and stool
samples {[}Table S10 \& S11{]}. Finally, these data suggest that the
most significantly increased OR genera for tissue was \emph{Lepttrichia}
while in stool it was \emph{Peptostreptococcus} {[}Table S10 \& S11{]}.

\textbf{\emph{Select Community Models can Recapitulate Whole Community
Models:}} Since specific genera increased OR for carcinoma over
diversity metrics we assessed whether the bacterial community was better
at classifying disease versus only a select group of genera. We selected
these genera based on their OR and P-value significance and used two
approaches to test this question. The first approach used genus level
data and tested for differences in AUC between all genera and selected
genera. A single study was used for training the model prior to testing
on all other studies and this was repeated for every study in the
meta-analysis. The second approach used OTU level data and tested for a
generalized decrease in the 10-fold cross validation (CV) model AUC
which is a common approach used to guard against over-fitting. This was
applied across study and the AUC of the all OTUs model was compared
against the model that used only OTUs that taxonomically classified to
selected genera.

For the first approach using the genera-based models, the training set
median AUC for model classification was similar for both the select
genera and full genera models, for both tissue and stool studies
{[}Figure S2-S3{]}. When analyzing the tests sets that were comprised of
genera data from other studies, both models had a similar ability to
detect individuals with adenomas or carcinomas, with the select genera
models performing better in some instances {[}Figure S4-S6{]}.
Conversely, the second approach that used OTU-based models showed a
slight decrease in median AUC between the full and select models, with
one exception to this generalization being the carcinoma models for
matched colon tissue {[}Figure 4 \& 5{]}.

@@

\textbf{Figure S1: Relative Risk for Adenoma or Carcinoma based on
Bacterial Community Metrics in Tissue.} A) Community-based relative risk
for adenoma. B) Community-based relative risk for carcinoma. Colors
represent the different variable regions used within the respective
study.

>>>>

\textbf{\emph{Mouth-Associated Genera are Associated with an Increased
OR of Tumor:}} Next, we asked if being higher than the median relative
abundance, for any specific genera, resulted in an altered OR for
adenoma or carcinoma, in stool and colon tissue, due to our previous
observations of small increases in OR using OTU richness and Shannon
diversity. To investigate this we analyzed all common genera across each
study, in colon tissue or stool, and assessed whether a relative
abundance higher than the median results in an increase or decrease in
OR. Mouth-associated genera were commonly found in the top 5 genera
associated with an increased OR of having an adenoma
(\emph{Porphyromonas} {[}Figure 3A{]} and \emph{Rothia} {[}Figure 3C{]})
and carcinoma (\emph{Fusobacterium}, \emph{Parvimonas},
\emph{Porphyromonas}, and \emph{Peptostreptococcus} {[}Figure 3B{]} and
\emph{Fusobacterium} and \emph{Parvimonas} {[}Figure 3D{]}) for both
stool and colon tissue samples. Conversely, genera commonly associated
with the gastrointestinal tract were correlated with a decreased OR for
both adenoma and carcinoma for both stool and colon tissue samples
{[}Figure 3{]}. Even though mouth-associated genera were identified
across disease stage, there was little direct overlap of the top 5
increased or decreased OR genera between both stages and sample site.

When observing ORs for adenoma between genera from stool or colon tissue
with a P-value less than 0.05 there was almost no overlap and when they
were similar the OR was in opposite directions (e.g. \emph{Lactococcus})
{[}Table S8 \& S9{]}. Many of the adenoma associated genera ORs with a
P-value under 0.05 for colon tissue are also highly prevalent in
contamination, specifically, \emph{Novosphingobium}, \emph{Pseudomonas},
and \emph{Achromobacter} {[}Figure 3 \& Table S8-S9{]}. For carcinoma
stage of disease, certain mouth-associated genera (\emph{Fusobacterium},
\emph{Parvimonas}) had an increased OR for both colon tissue and stool
samples {[}Table S10 \& S11{]}. The genera with the highest increased OR
for carcinoma in tissue was \emph{Lepttrichia} while in stool it was
\emph{Peptostreptococcus} {[}Table S10 \& S11{]}.

\textbf{\emph{Select Community Models can Recapitulate Whole Community
Models:}} Since specific genera increased the OR for carcinoma over
diversity metrics we assessed whether the bacterial community was better
at classifying disease versus only a select group of genera. We selected
these genera based on their OR and P-value significance and used two
approaches to test this question. The first approach used genus level
data and tested for differences in AUC between all genera and selected
genera. A single study was used for training the model prior to testing
on all other studies and this was repeated for every study in the
meta-analysis. The second approach used OTU level data and tested for a
generalized decrease in the 10-fold cross validation (CV) model AUC
which is a common approach used to guard against over-fitting. This was
applied across study and the AUC of the all OTUs model was compared
against the model that used only OTUs that taxonomically classified to
selected genera.

For the first approach using the genera-based models, the training set
median AUC for model classification was similar for both the full and
select genera models, for both tissue and stool studies {[}Figure
S2-S3{]}. When analyzing the tests sets that were comprised of genera
data from other studies, both models had a similar ability to detect
individuals with adenomas or carcinomas, with the select genera models
performing better in some instances {[}Figure S4-S6{]}. Conversely, the
second approach that used OTU-based models showed a slight decrease in
median AUC between the full and select models {[}Figure 4 \& 5{]}.

@@

\textbf{Figure S1: Odds Ratio for Adenoma or Carcinoma based on
Bacterial Community Metrics in Tissue.} A) Community-based odds ratio
for adenoma. B) Community-based odds ratio for carcinoma. Colors
represent the different variable regions used within the respective
study.
